Modeling the near-wake of a vertical-axis cross-flow turbine with 2-D and 3-D RANS

### 2016-03-19 paper.tex
Copy stuff from CFD thesis chapter up to results

\section{Numerical models}

The flow field was modeled using the Reynolds-averaged Navier--Stokes equations,
employing two different turbulence models---Menter's $k$--$\omega$ SST
\cite{Menter1994} and the Spalart--Allmaras (SA) one equation model
\cite{Spalart1992}. The SST model was chosen due to its prominence in the
literature for simulating separating flows, which we assumed to be present in
the current problem in the form of dynamic stall. The SA model was shown by
Ferreira et al. \cite{Ferreira2007} to match experimental particle image
velocimetry (PIV) results for a CFT in dynamic stall, though this was a somewhat
low Reynolds number case ($5 \times 10^4$). Further justification for using the
SA model for this case comes from Crivellini and D'Alessandro
\cite{Crivellini2014}, where they successfully modeled the laminar separation
bubble and subsequent boundary layer transition to turbulence at Reynolds
numbers similar to ours.

@@

The computational domain is a rectangular volume 3.66 m long, 3.66 m wide, and
2.44 m tall (for 3-D), with the turbine located 1.52 m from the inlet, and
centered vertically with a vertical axis. with a cylindrical sliding mesh
interface, where the turbine zone rotates at a mean tip speed ratio
$\lambda=1.9$ with a sinusoidal oscillation at the blade passage
frequency---with and amplitude of 0.19 and the first peak at 0.7 radians---to
mimic the experimental data. The 2-D mesh overview is shown in
Figure~\ref{fig:mesh} and the blade mesh is shown in
Figure~\ref{fig:blade-mesh}.

@@

3-D computer aided design (CAD) files of the turbine are available from
\cite{Bachant2014-RVAT-CAD}.

\begin{figure}[ht]    
    \includegraphics[width=0.8\textwidth]{figures/2d_mesh}

@@

    \label{fig:mesh}
\end{figure}

\begin{figure}[ht]
    \includegraphics[width=0.7\textwidth]{figures/2D_blade_mesh_closeup}

@@

Simulations were run using the \textit{pimpleDyMFoam} solver from the
open-source finite volume CFD package \textit{OpenFOAM}, version 2.3.x.
\textit{pimpleDyMFoam} uses a hybrid PISO-SIMPLE algorithm for pressure-velocity
coupling and is compatible with dynamic meshes.

@@

Initial and boundary conditions were set to match those of the tow tank as well
as possible. The velocity at the bottom and side walls was fixed to 1 m/s to
match the tow tank case, while the top boundary condition was a slip velocity
condition. Note that in 2-D the top and bottom boundary conditions are
``empty,'' which is an OpenFOAM convention.

@@

The $k$--$\omega$ SST and Spalart--Allmaras RANS model cases were verified for
convergence of the turbine mean power coefficient with respect to grid spacing
and timestep. The grid topology was fixed, but the number of cells per unit
length were scaled proportionally, maintaining the same background mesh cell
aspect ratio. Results for this parameter sweep are shown in
Figure~\ref{fig:verification}, from which the final number of streamwise
grid points $N_x = 70$ was chosen. This corresponds to a total cell count of
approximately $5 \times 10^4$ for the 2-D cases and 16 million for the 3-D case.

    \includegraphics[width=0.9\textwidth]{figures/verification}

@@

    \label{fig:verification}
\end{figure}

Time step dependence was studied using the $N_x=70$ grid, the results from which
are shown in Figure~\ref{fig:verification}. It was seen that the
Spalart--Allmaras model converged well with decreasing time step, leading to a
final time step of 0.001 s. The results from the SST model show a local minimum
at $\Delta t = 0.002$ s, with some divergence for smaller time steps. The local
minimum was chosen as the final time step to run the simulations. Note that the
SST model's convergence behavior may be due to its specific implementation in
\textit{OpenFOAM}, and not indicative of the nature of the model equations.
Verification studies for CFTs with this level of detail in the literature are
not common, though the final time step is comparable to others
\cite{Balduzzi2016}.

>>>>

\section{Numerical setup}

In this study the UNH-RVAT baseline tow tank experiment was simulated using a
mean rotor tip speed ratio $\lambda=1.9$, which corresponds to the maximum
measured power coefficient. The tow speed or free stream velocity $U_\infty=1.0$
m/s gives a turbine diameter Reynolds number $Re_D \approx 10^6$, which
corresponds to the $Re$-independent state for both performance and near-wake
characteristics, as determined from previous experimental measurements
\cite{Bachant2014, Bachant2016-Energies}.

The turbines CAD geometry was prepared or ``cleaned'' for CFD by removing
details determined to be unnecessary, e.g., screw heads and axial shaft grooves,
which would complicate the meshing process and contribute very little to the
overall loading or flow modulation. A drawing of the physically and numerically
modeled geometries is presented in Figure~\ref{fig:cfd-cad}.

    \includegraphics[clip, trim=0 1in 0 1in, width=0.8\textwidth]{BR-CFD_CAD}

    \label{fig:cfd-cad}
\end{figure}

To close the RANS equations, two different turbulence models were
used---Menter's $k$--$\omega$ SST \cite{Menter1994} and the Spalart--Allmaras
(SA) one equation model \cite{Spalart1992}. Both closures use the eddy-viscosity
approach---the SA employing a single additional scalar transport equation and
the SST two. The SST model was chosen due to its prominence in the literature
for simulating separating flows, which we assumed to be present in the current
problem in the form of dynamic stall. The SA model was shown by Ferreira
et al.~\cite{Ferreira2007} to match experimental particle image
velocimetry (PIV) results for a CFT in dynamic stall, though this was a somewhat
low Reynolds number case ($5 \times 10^4$). Further justification for using the
SA model for this case comes from Crivellini and D'Alessandro
\cite{Crivellini2014}, where they successfully modeled the laminar separation
bubble and subsequent boundary layer transition to turbulence at Reynolds
numbers similar to those investigated here.

@@

The computational domain was a rectangular volume 3.66 m long, 3.66 m wide, and
2.44 m tall (for 3-D), with the turbine located 1.52 m from the inlet, and
centered vertically with a vertical axis, designed to match the tow tank
dimensions for comparison with previous experiments. The rotor geometry was
located at the center of a cylindrical sliding mesh interface, which rotated at
a mean tip speed ratio $\lambda=1.9$ with a sinusoidal oscillation at the blade
passage frequency---with and amplitude of 0.19 and the first peak at 1.4
radians---to mimic the slight deviation from the mean tip speed ratio observed
in the experiments. The 2-D mesh overview is shown in
Figure~\ref{fig:2d-br-mesh} and the blade mesh detail is shown in
Figure~\ref{fig:blade-mesh}.

@@

    \includegraphics[width=0.8\textwidth]{BR-CFD_2D_mesh}

@@

    \label{fig:2d-br-mesh}
\end{figure}

    \includegraphics[width=0.7\textwidth]{BR-CFD_2D_blade_mesh_closeup}

@@

Simulations were run using the \textit{pimpleDyMFoam} solver from the
open-source finite volume CFD package \textit{OpenFOAM}, version 2.3.x.
\textit{pimpleDyMFoam} uses a hybrid PISO-SIMPLE algorithm for pressure-velocity
coupling and is compatible with dynamic meshes. An Euler scheme was used to
advance the simulation forward in time. The case files required to replicate the
simulations are available from \cite{Bachant2016-UNH-RVAT-2D-OpenFOAM-SST,
    Bachant2016-UNH-RVAT-2D-OpenFOAM-SA, Bachant2016-UNH-RVAT-3D-OpenFOAM-SST,
    Bachant2016-UNH-RVAT-3D-OpenFOAM-SA}.

@@

Initial and boundary conditions were set to match those of the tow tank as well
as possible. The velocity at the inlet, bottom, and side walls was fixed to 1
m/s to match the tow tank case, while the top boundary condition was a slip
velocity condition. Pressure was held fixed at the outlet. Note that in 2-D the
top and bottom boundary conditions are ``empty,'' which is an OpenFOAM
convention to indicate two-dimensionality.

Since the mesh was only refined next to the blade surfaces in order to resolve
the boundary layer profile, no wall functions were used. However, wall functions
were used for the other turbine surfaces, i.e., struts, hub, and shaft, and for
the tank sidewalls, bottom, and top.

@@

Both the $k$--$\omega$ SST and Spalart--Allmaras RANS model cases were verified
for convergence of the turbine mean power coefficient with respect to grid
spacing and time step using 2-D domains. The grid topology was fixed, but the
number of cells per domain length was scaled proportionally, maintaining the
same background mesh cell aspect ratio. Results for this parameter sweep are
shown in Figure~\ref{fig:2d-br-verification}, from which the final number of
streamwise grid points $N_x = 70$ was chosen. This corresponds to a total cell
count of approximately $5 \times 10^4$ for the 2-D cases and 16 million for the
3-D case.

    \includegraphics[width=0.9\textwidth]{BR-CFD_verification}

@@

    \label{fig:2d-br-verification}
\end{figure}

Time step dependence was evaluated using the 2-D $N_x=70$ grid, the results from
which are shown in Figure~\ref{fig:2d-br-verification}. It was seen that the
Spalart--Allmaras model converged well with decreasing time step, leading to a
final time step of 0.001 s. The results from the SST model show a local minimum
at $\Delta t = 0.002$ s, with some divergence for smaller time steps. The local
minimum was chosen as the final time step to run the simulations. Note that the
SST model's convergence behavior may be due to its specific implementation in
\textit{OpenFOAM}, and not indicative of the nature of the model equations.
Verification studies for CFTs with this level of detail in the literature are
not common, though the final time step is comparable to others
\cite{Balduzzi2016}.
